---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidypolars

---

:warning: If you're looking for the Python package "tidypolars" by Mark Fairbanks,
you're on the wrong repo. The right one is here: [markfairbanks/tidypolars](https://github.com/markfairbanks/tidypolars) :warning:

---


`polars` (both the Rust source and the R implementation) are amazing packages.
I won't argue here for the interest of using `polars`, there are already a lot 
of resources on [its website](https://rpolars.github.io/).

One characteristic of `polars` is that its syntax is 1) extremely verbose, and 
2) very close to the `pandas` syntax in Python. While this makes it quite easy
to read, it is **yet another syntax to learn** for R users that are accustomed so
far to either base R, `data.table` or the `tidyverse`.

The objective of `tidypolars` is to **provide functions that are very close to
the `tidyverse` ones** but that call the `polars` functions under the hood so that
we don't lose anything of its capacities. Morevoer, the objective is to keep
`tidypolars` **dependency-free** with the exception of `polars` itself (which
has no dependencies).


## General syntax

Overall, you only need to add `pl_` as a prefix to the `tidyverse` function you're
used to. For example, `dplyr::mutate()` modifies classic `R` dataframes, and 
`tidypolars::pl_mutate()` modifies `polars`' `DataFrame`s and `LazyFrame`s.

## Example

```{r}
library(polars)
library(tidypolars)

pl_test <- pl$DataFrame(iris)
pl_test

pl_test |> 
  pl_filter(Species == "setosa") |> 
  pl_arrange(Sepal.Width, -Sepal.Length)

pl_test |> 
  pl_mutate(
    Sepal.Total = Sepal.Length + Sepal.Width,
    Petal.Total = Petal.Length + Petal.Width
  ) |> 
  pl_select(ends_with("Total"))
```


# FAQ

## Is `tidypolars` slower than `polars`?

No, or just marginally. The objective of `tidypolars` is *not* to modify the
data, simply to translate the `tidyverse` syntax to `polars` syntax. `polars`
is still in charge of doing all the data manipulations under the hood.

Therefore, there might be minor overhead because we still need to parse the 
expressions and rewrite them in `polars` syntax but this should be extremely 
marginal.

## Am I stuck with `tidypolars`?

No, as said above, `tidypolars` just changes one syntax to another but it doesn't
touch the data itself. So if for some reason you want to go back to a "raw" 
`polars` syntax later in your code, you're free to do so because `tidypolars`
will always return `DataFrame`s, `LazyFrame`s or `Series`.

## Do I still need to load `polars`?

Yes, because `tidypolars` doesn't provide any functions to create `polars`
`DataFrame` or `LazyFrame`, or to read data. You'll still need to use `polars`
for this.


## Can I see some benchmarks?

Sure but take them with a grain of salt: these small benchmarks may not be 
representative of real-life scenarios and don't necessarily use the full capacities
of other packages (e.g keyed `data.table`s). You should refer to [DuckDB benchmarks](https://duckdblabs.github.io/db-benchmark/) for more serious ones.

```{r}
library(polars)
library(tidypolars)
library(dplyr)
library(data.table)

test <- data.frame(
  grp = sample(letters, 1e7, TRUE),
  val1 = sample(1:1000, 1e7, TRUE),
  val2 = sample(1:1000, 1e7, TRUE)
)

pl_test <- pl$DataFrame(test)
dt_test <- as.data.table(test)

bench::mark(
  polars = pl_test$
    groupby("grp")$
    agg(
      pl$col('val1')$mean()$alias('x'), 
      pl$col('val2')$sum()$alias('y')
    ),
  tidypolars = pl_test |> 
    pl_group_by(grp) |> 
    pl_summarize(
      x = mean(val1),
      y = sum(val2)
    ),
  dplyr = test |> 
    group_by(grp) |> 
    summarize(
      x = mean(val1),
      y = sum(val2)
    ),
  data.table = dt_test[, .(x = mean(val1), y = sum(val2)), by = grp],
  check = FALSE
)
```

