---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidypolars

<!-- badges: start -->
[![check](https://github.com/etiennebacher/tidypolars/actions/workflows/check.yml/badge.svg)](https://github.com/etiennebacher/tidypolars/actions/workflows/check.yml)
[![tidypolars status badge](https://etiennebacher.r-universe.dev/badges/tidypolars)](https://etiennebacher.r-universe.dev/tidypolars)
[![Codecov test coverage](https://codecov.io/gh/etiennebacher/tidypolars/branch/main/graph/badge.svg)](https://app.codecov.io/gh/etiennebacher/tidypolars?branch=main)
<!-- badges: end -->


---

:warning: If you're looking for the Python package "tidypolars",
you're on the wrong repo. The right one is here: [markfairbanks/tidypolars](https://github.com/markfairbanks/tidypolars) :warning:

---

* [Motivation](#motivation)
* [Installation](#installation)
* [Example](#example)


## Motivation

`polars` is a DataFrame library written in Rust and with bindings in several
languages, [including R](https://rpolars.github.io/). I won't argue here for the
interest of using `polars`, there are already a lot of resources on [its
website](https://www.pola.rs/).

The R package `polars` was made to mimic very closely the original Python syntax,
which is quite verbose. While this makes it easy to read, it is **yet another 
syntax to learn** for R users that are accustomed so far to either base R, 
`data.table` or the `tidyverse`.

The objective of `tidypolars` is to **provide the power and speed of `polars` 
while keeping the `tidyverse` syntax**. 


## Installation

`tidypolars` is built on `polars`, which is not available on CRAN. This means 
that `tidypolars` also can't be on CRAN. However, you can install it from 
R-universe.

### Windows or macOS

```{r eval=FALSE}
install.packages(
  'tidypolars', 
  repos = c('https://etiennebacher.r-universe.dev', getOption("repos"))
)
```

### Linux

```{r eval=FALSE}
install.packages(
  'tidypolars', 
  repos = c('https://etiennebacher.r-universe.dev/bin/linux/jammy/4.3', getOption("repos"))
)
```


## Example

Suppose that you already have some code that uses `dplyr`:

```{r}
library(dplyr, warn.conflicts = FALSE)

iris |> 
  select(starts_with(c("Sep", "Pet"))) |> 
  mutate(
    petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
  ) |> 
  filter(between(Sepal.Length, 4.5, 5.5)) |> 
  head()
```

With `tidypolars`, you can provide a Polars `DataFrame` or `LazyFrame` and keep 
the exact same code:

```{r}
library(tidypolars)

iris |> 
  as_polars() |> 
  select(starts_with(c("Sep", "Pet"))) |> 
  mutate(
    petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
  ) |> 
  filter(between(Sepal.Length, 4.5, 5.5)) |> 
  head()
```

If you're used to the `tidyverse` functions and syntax, this will feel much 
easier to read than the pure `polars` syntax:

```{r}
library(polars)

# polars syntax
pl$DataFrame(iris)$
  select(c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))$
  with_columns(
    pl$when(
      (pl$col("Petal.Length") / pl$col("Petal.Width") > 3)
    )$then(pl$lit("long"))$
      otherwise(pl$lit("large"))$
      alias("petal_type")
  )$
  filter(pl$col("Sepal.Length")$is_between(4.5, 5.5))$
  head(6)
```

Since most of the work is rewriting `tidyverse` code into `polars` syntax, 
`tidypolars` and `polars` have very similar performance.

<details>
<summary>Click to see a small benchmark</summary>

```{r}
large_iris <- data.table::rbindlist(rep(list(iris), 50000))
large_iris_pl <- as_polars(large_iris, lazy = TRUE)

bench::mark(
  polars = {
    large_iris_pl$
      select(c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))$
      with_columns(
        pl$when(
          (pl$col("Petal.Length") / pl$col("Petal.Width") > 3)
        )$then(pl$lit("long"))$
          otherwise(pl$lit("large"))$
          alias("petal_type")
      )$
      filter(pl$col("Sepal.Length")$is_between(4.5, 5.5))$
      collect()
  },
  tidypolars = {
    large_iris_pl |>
      select(starts_with(c("Sep", "Pet"))) |>
      mutate(
        petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
      ) |> 
      filter(between(Sepal.Length, 4.5, 5.5)) |> 
      collect()
  },
  dplyr = {
    large_iris |>
      select(starts_with(c("Sep", "Pet"))) |>
      mutate(
        petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
      ) |>
      filter(between(Sepal.Length, 4.5, 5.5))
  },
  check = FALSE,
  iterations = 10
)

# NOTE: do NOT take the "mem_alloc" results into account.
# `bench::mark()` doesn't report the accurate memory usage for packages calling
# Rust code.
```
</details>


