---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidypolars

---

:warning: If you're looking for the Python package "tidypolars",
you're on the wrong repo. The right one is here: [markfairbanks/tidypolars](https://github.com/markfairbanks/tidypolars) :warning:

---

* [Motivation](#motivation) 
* [General syntax](#general-syntax) 
* [Example](#example)


## Motivation


`polars` (both the Rust source and the R implementation) are amazing packages.
I won't argue here for the interest of using `polars`, there are already a lot 
of resources on [its website](https://rpolars.github.io/).

One characteristic of `polars` is that its syntax is 1) extremely verbose, and 
2) very close to the `pandas` syntax in Python. While this makes it quite easy
to read, it is **yet another syntax to learn** for R users that are accustomed so
far to either base R, `data.table` or the `tidyverse`.

The objective of `tidypolars` is to **provide functions that are very close to
the `tidyverse` ones** but that call the `polars` functions under the hood so that
we don't lose anything of its capacities. Moreover, the objective is to keep
`tidypolars` **dependency-free** with the exception of `polars` itself (which
has no dependencies).


## General syntax

Overall, you only need to add `pl_` as a prefix to the `tidyverse` function you're
used to. For example, `dplyr::mutate()` modifies classic `R` dataframes, and 
`tidypolars::pl_mutate()` modifies `polars`' `DataFrame`s and `LazyFrame`s.

## Example

`tidypolars` makes it easier for R users to use `polars`:

```{r}
library(polars)
library(tidypolars)

# polars syntax
pl$DataFrame(iris)$
  select(c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))$
  with_columns(
    pl$when(
      (pl$col("Petal.Length") / pl$col("Petal.Width") > 3)
    )$then("long")$
      otherwise("large")$
      alias("petal_type")
  )

# tidypolars syntax
iris |> 
  as_polars() |> 
  pl_select(starts_with("Sep", "Pet")) |> 
  pl_mutate(
    petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
  )
```

And it's still as fast as the original `polars` syntax:

```{r}
library(dplyr, warn.conflicts = FALSE)

large_iris <- data.table::rbindlist(rep(list(iris), 100000))

bench::mark(
  polars = {
    pl$DataFrame(large_iris)$
      select(c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))$
      with_columns(
        pl$when(
          (pl$col("Petal.Length") / pl$col("Petal.Width") > 3)
        )$then("long")$
          otherwise("large")$
          alias("petal_type")
      )
  }, 
  tidypolars = {
    large_iris |> 
      as_polars() |> 
      pl_select(starts_with("Sep", "Pet")) |> 
      pl_mutate(
        petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
      )
  },
  dplyr = {
    large_iris |> 
      select(starts_with(c("Sep", "Pet"))) |> 
      mutate(
        petal_type = ifelse((Petal.Length / Petal.Width) > 3, "long", "large")
      )
  },
  check = FALSE,
  iterations = 10
)
```
