---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidypolars

<!-- badges: start -->
<!-- badges: end -->

`polars` (both the Rust source and the R implementation) are amazing packages.
I won't argue here for the interest of using `polars`, there are already a lot 
of resources of [its website]().

One characteristic of `polars` is that its syntax is 1) extremely verbose, and 
2) very close to the `pandas` syntax in Python. While this makes it quite easy
to read, it is **yet another syntax to learn** for R users that are accustomed so
far to either base R, `data.table` or the `tidyverse`.

The objective of `tidypolars` is to **provide functions that are very close to
the `tidyverse` ones** but that call the `polars` functions under the hood so that
we don't lose anything of its capacities. Morevoer, the objective is to keep
`tidypolars` **dependency-free** with the exception of `polars` itself (which
has no dependencies).


## General syntax

Overall, you only need to add `pl_` as a prefix to the `tidyverse` function you're
used to. For example, `dplyr::mutate()` modifies classic `R` dataframes, and 
`tidypolars::pl_mutate()` modifies `polars`' `DataFrame`s and `LazyFrame`s.

## Example

```{r}
library(polars)
library(tidypolars)

pl_test <- pl$DataFrame(iris)
pl_test

pl_test |> 
  pl_filter(Species == "setosa") |> 
  pl_arrange(Sepal.Width, -Sepal.Length)

pl_test |> 
  pl_mutate(
    Sepal.Total = Sepal.Length + Sepal.Width,
    Petal.Total = Petal.Length + Petal.Width
  ) |> 
  pl_select(ends_with("Total"))
```


# FAQ

## Is `tidypolars` slower than `polars`?

No, or just marginally. The objective of `tidypolars` is *not* to modify the
data, simply to translate the `tidyverse` syntax to `polars` syntax. `polars`
is still in charge of doing all the data manipulations under the hood.

Therefore, there might be minor overhead because we still need to parse the 
expressions and rewrite them in `polars` syntax but this should be extremely 
marginal.
