---
title: "FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidypolars)
```


# FAQ

## Is `tidypolars` slower than `polars`?

No, or just marginally. The objective of `tidypolars` is *not* to modify the
data, simply to translate the `tidyverse` syntax to `polars` syntax. `polars`
is still in charge of doing all the data manipulations under the hood.

Therefore, there might be minor overhead because we still need to parse the 
expressions and rewrite them in `polars` syntax but this should be extremely 
marginal.

## Am I stuck with `tidypolars`?

No, as said above, `tidypolars` just changes one syntax to another but it doesn't
touch the data itself. So if for some reason you want to go back to a "raw" 
`polars` syntax later in your code, you're free to do so because `tidypolars`
will always return `DataFrame`s, `LazyFrame`s or `Series`.

## Do I still need to load `polars`?

Yes, because `tidypolars` doesn't provide any functions to create `polars`
`DataFrame` or `LazyFrame`, or to read data. You'll still need to use `polars`
for this.


## Can I see some benchmarks?

Sure but take them with a grain of salt: these small benchmarks may not be 
representative of real-life scenarios and don't necessarily use the full capacities
of other packages (e.g keyed `data.table`s). You should refer to [DuckDB benchmarks](https://duckdblabs.github.io/db-benchmark/) for more serious ones.

```{r}
library(polars)
library(tidypolars)
library(dplyr, warn.conflicts = FALSE)
library(data.table, warn.conflicts = FALSE)

test <- data.frame(
  grp = sample(letters, 1e7, TRUE),
  val1 = sample(1:1000, 1e7, TRUE),
  val2 = sample(1:1000, 1e7, TRUE)
)

pl_test <- pl$DataFrame(test)
dt_test <- as.data.table(test)

bench::mark(
  polars = pl_test$
    groupby("grp")$
    agg(
      pl$col('val1')$mean()$alias('x'), 
      pl$col('val2')$sum()$alias('y')
    ),
  tidypolars = pl_test |> 
    pl_group_by(grp) |> 
    pl_summarize(
      x = mean(val1),
      y = sum(val2)
    ),
  dplyr = test |> 
    group_by(grp) |> 
    summarize(
      x = mean(val1),
      y = sum(val2)
    ),
  data.table = dt_test[, .(x = mean(val1), y = sum(val2)), by = grp],
  check = FALSE,
  iterations = 15
)

bench::mark(
  polars = pl_test$
    filter(pl$col("grp") == "a" | pl$col("grp") == "b"),
  tidypolars = pl_test |> 
    pl_filter(grp == "a" | grp == "b"),
  dplyr = test |> 
    filter(grp %in% c("a", "b")),
  data.table = dt_test[grp %chin% c("a", "b")],
  check = FALSE,
  iterations = 15
)
```
