---
title: "R and Polars expressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R and Polars expressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


When we use the `tidyverse`, we use R expressions in mainly three places: `filter()`,
`mutate()`, and `summarize()`.

```{r echo=FALSE}
library(dplyr, warn.conflicts = FALSE)
```


```{r eval=FALSE}
library(dplyr, warn.conflicts = FALSE)

filter(mtcars, am + gear > carb)
mutate(mtcars, x = (qsec - mean(qsec) / sd(qsec)))
mtcars |> 
  group_by(cyl) |> 
  summarize(x = mean(qsec) / sd(qsec))
```

This is very convenient but creates a challenge for `tidypolars`. Indeed, while
it is possible to pass R functions directly to a Polars Data/LazyFrame, it is
**strongly discouraged** to do so because it doesn't take advantage of Polars
optimizations. 

Indeed, Polars comes with dozens of built-in functions for maths (`median`, `var`, `arccos`, ...), string manipulation (`n_chars`, `starts`, ...), and date-time (`hour`, `quarter`, `ordinal_day`, ...). All of these functions are optimized internally and are ran in parallel under the hood, which will not be the case if we pass R functions.

However, using these Polars expressions would imply that we need to learn these new functions and this new syntax. To avoid doing that, `tidypolars` will automatically translate R expressions into Polars ones. Basically, **you can keep writing R expressions in most situations**, and they will automatically be translated to Polars syntax. 

However, there are some situations where this might not work, so this vignette explains the process and the limitations.

# How does `tidypolars` translate R expressions into Polars expressions?

When `tidypolars` receives an expression, it runs a function `translate()` several times until all components are translated to their Polars equivalent. There are four possible components: single values, column names, external objects, and functions.


## Single values, column names, and external objects

If you pass a single value, like `x = 1` or `x = "a"`, it is wrapped into `pl$lit()`. This is also the case for external objects with the difference that these are evaluated before they are wrap into `pl$lit()`.

Column names, like `x = mpg`, are wrapped into `pl$col()`. 

```
x = "a"     -> x = pl$lit("a")
x = letters -> x = pl$lit(letters)
x = mpg     -> x = pl$col("mpg")
```

## Functions


Functions are split into two categories: built-in functions (i.e functions provided by base R or by other packages), and user-defined functions (UDF) that are written by the user (you). 


### Built-in functions

In the first case, `tidypolars` checks the function name and whether there's an equivalent function in Polars. For example, the R function `sd(x, na.rm = TRUE)` is converted to `std(x, na.rm = TRUE)`. Since R and Polars functions don't share the same name, this check relies on a custom list containing all equivalencies between R and Polars functions. You can see the list of supported R functions at the bottom of this vignette.

Here, replacing R's `sd()` functions with `std()` is not enough because the argument `x` (which usually is a column of the dataset) is not in the right format to be used by Polars. Therefore, `tidypolars` calls `translate()` a second time on the inside of the function.  

We now have `std(pl$col("x"))`^[Not exactly because `tidypolars` actually calls an internal function `pl_std()` that contains the right Polars syntax, but this is the main idea.]. To end this example, we need to see what happens to additional arguments. You can see that we didn't modify the `na.rm = TRUE` for now. This is because Polars doesn't have this argument in `std()` (it automatically drops the missing values). Internally, `tidypolars` checks whether additional arguments are accepted and throws a message if this is not the case:

```{r}
library(tidypolars)
library(polars)

mtcars |> 
  as_polars() |> 
  pl_mutate(x = sd(mpg, na.rm = TRUE))
```


### User-defined functions

User-defined functions (UDF) are more challenging. Indeed, it is technically possible to inspect the code inside a UDF, but rewriting it to match Polars syntax would be extremely complicated. In this situation, you will have to rewrite your custom function using Polars syntax so that it returns a Polars expression. For example, we could make a function to standardize a column like this:


```{r}
pl_standardize <- function(x) {
  (x - x$mean()) / x$std()
}
```

Remember that the column name used as `x` will end up wrapped into `pl$col()`, so if you want to check if your function returns a Polars expression, you have to provide a `pl$col()` call:

```{r}
pl_standardize(pl$col("mpg"))
```

This function correctly returns a Polars expression, so we can now use it like any other function:

```{r}
mtcars |> 
  as_polars() |> 
  pl_mutate(x = pl_standardize(mpg))
```


### Special case: `across()`

[`across()`](https://dplyr.tidyverse.org/reference/across.html) is a very useful function that applies a function (or a list of functions) to a selection of columns. It also accepts built-in functions and UDFs.

```{r}
mtcars |> 
  as_polars() |> 
  pl_mutate(
    across(
      .cols = contains("a"),
      list(mean = mean, stand = pl_standardize, ~ sd(.x))
    )
  )
```

However, it will also error if the UDF you pass doesn't return a Polars expression. Similarly, if you want to pass an anonymous function, it has to return a Polars expression:

```{r error=TRUE}
mtcars |> 
  as_polars() |> 
  pl_mutate(
    across(
      .cols = contains("a"),
      .fns = list(
        mean = mean,  
        function(x) {
           (x - mean(x)) / sd(x)
        },
        ~ sd(.x)
      )
    )
  )
```


## List of base R and `tidyverse` functions supported by `tidypolars` 


```{r echo=FALSE, message=FALSE}
library(dplyr)
library(knitr)
out <- tribble(
  ~Package, ~Function,
  "`base`", "`abs`",
  "`base`", "`acos`", 
  "`base`", "`acosh`", 
  "`base`", "`asin`",
  "`base`", "`asinh`", 
  "`base`", "`atan`", 
  "`base`", "`atanh`",
  "`base`", "`ceiling`",
  "`base`", "`cos`", 
  "`base`", "`cosh`",
  "`base`", "`cummin`",
  "`base`", "`cumsum`", 
  "`base`", "`exp`", 
  "`base`", "`floor`",
  "`base`", "`ifelse`",
  "`stats` / `dplyr`", "`lag`",
  "`base`", "`log`", 
  "`base`", "`log10`",
  "`base`", "`max`", 
  "`base`", "`mean`", 
  "`stats`", "`median`", 
  "`base`", "`min`",
  "`base`", "`nchar`",
  "`base`", "`paste0`",
  "`base`", "`rank`",
  "`base`", "`round`",
  "`base`", "`sin`", 
  "`base`", "`sinh`", 
  "`base`", "`sort`",  
  "`base`", "`sqrt`", 
  "`stats`", "`sd`", 
  "`base`", "`tan`", 
  "`base`", "`tanh`",
  "`base`", "`tolower`",
  "`base`", "`toupper`",
  "`stats`", "`var`",
  "`dplyr`", "`between`",
  "`dplyr`", "`case_match`",
  "`dplyr`", "`case_when`",
  "`dplyr`", "`coalesce`",
  "`dplyr`", "`first`",
  "`dplyr`", "`if_else`",
  "`dplyr`", "`last`",
  "`stringr`", "`str_count`",
  "`stringr`", "`str_ends`",
  "`stringr`", "`str_extract`",
  "`stringr`", "`str_length`",
  "`stringr`", "`str_remove`",
  "`stringr`", "`str_remove_all`",
  "`stringr`", "`str_replace`",
  "`stringr`", "`str_replace_all`",
  "`stringr`", "`str_starts`",
  "`stringr`", "`str_sub`",
  "`stringr`", "`str_to_lower`",
  "`stringr`", "`str_to_upper`",
) |> 
  arrange(Package)

rows <- seq_len(nrow(out) %/% 2)
kable(list(out[rows,], matrix(numeric(), nrow=0, ncol=1),
           out[-rows, ])) 
```

