---
title: "Parsing expressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parsing expressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Some functions are quite easy to translate in Polars syntax, such as `distinct()`
(which calls `$unique()` under the hood). Things are more challenging when we
try to convert classic R expressions to Polars expressions to use them in 
`mutate()` or `summarize()`. This article walks you through the process of 
converting these expressions, so that you can understand the strengths and 
weaknesses of `tidypolars`.


## Why should we use Polars expressions?

Polars provides a way to evaluate custom functions using `$apply()` or `$map()`.
However, **this is strongly discouraged** and people should use Polars' numerous
built-in functions, that are optimized and parallelized (most of the time) 
internally.

Therefore, there seems to be a conflict between the two objectives of `tidypolars`:

* provide Polars performance
* keep using R and `tidyverse` syntax


## How does `tidypolars` work with expressions? 

We need to parse expressions that R users provide and convert them to Polars 
syntax. To do so, `tidypolars` works in several steps.

The main tools in this rewriting process are `parse()` and `getParseData()`. For
example, suppose we have this string that contains an expression that a user wants
to pass in `pl_mutate()`:

```{r}
expr <- "mean(Petal.Length) / sum(Sepal.Width)"
```

Using `parse()` and `getParseData()`, we can represent this as a table with each
element:

```{r}
parsed <- parse(text = expr, keep.source = TRUE)
p_d <- utils::getParseData(parsed)
p_d
```

In particular, we are interested in the columns `token` and `text`. To get variable
names and replace them with `pl$col(varname)`, we keep only `token == SYMBOL` and 
then we check which `text` values match the column names of our data. 

We can also extract function calls with `token == SYMBOL_FUNCTION_CALL`. For each
function call, we check whether there's a Polars equivalent. If there is 
one, we prefix the function call with `pl_`. For example, `mean(x)` will be 
transformed as `pl_mean(x)`, which is defined as a very short function that
returns the Polars expression `x$mean()`.

Finally, one last challenge is that R functions and their Polars equivalent 
do not necessarily have the same arguments. For example, `mean()` has an argument
`na.rm` to drop missing values. In Polars, this is done automatically. Overall,
when some elements of the R function are not used by Polars, it will generate
a message of the form:

```r
When the dataset is a Polars DataFrame or LazyFrame, `mean()` only needs 
one argument. Additional arguments will not be used. 

```

Now, you know how `tidypolars` converts R expressions into Polars expressions.


## Drawback

A drawback of this approach is that `tidypolars` can only work with the functions
it was programmed to know. For example, it recognizes `ifelse()` but if this
`ifelse()` is buried in a custom function that you defined elsewhere,
it is not possible to convert it to Polars syntax so you will lose the 
performance gains here.


## List of base R and `tidyverse` functions supported by `tidypolars` 


```{r echo=FALSE, message=FALSE}
library(dplyr)
library(knitr)
out <- tribble(
  ~Package, ~Function,
  "`base`", "`abs`",
  "`base`", "`acos`", 
  "`base`", "`acosh`", 
  "`base`", "`asin`",
  "`base`", "`asinh`", 
  "`base`", "`atan`", 
  "`base`", "`atanh`",
  "`base`", "`ceiling`",
  "`base`", "`cos`", 
  "`base`", "`cosh`",
  "`base`", "`cummin`",
  "`base`", "`cumsum`", 
  "`base`", "`exp`", 
  "`base`", "`floor`",
  "`base`", "`ifelse`",
  "`stats` / `dplyr`", "`lag`",
  "`base`", "`log`", 
  "`base`", "`log10`",
  "`base`", "`max`", 
  "`base`", "`mean`", 
  "`stats`", "`median`", 
  "`base`", "`min`",
  "`base`", "`nchar`",
  "`base`", "`paste0`",
  "`base`", "`rank`",
  "`base`", "`round`",
  "`base`", "`sin`", 
  "`base`", "`sinh`", 
  "`base`", "`sort`",  
  "`base`", "`sqrt`", 
  "`stats`", "`sd`", 
  "`base`", "`tan`", 
  "`base`", "`tanh`",
  "`base`", "`tolower`",
  "`base`", "`toupper`",
  "`stats`", "`var`",
  "`dplyr`", "`between`",
  "`dplyr`", "`case_match`",
  "`dplyr`", "`case_when`",
  "`dplyr`", "`coalesce`",
  "`dplyr`", "`first`",
  "`dplyr`", "`if_else`",
  "`dplyr`", "`last`",
  "`stringr`", "`str_count`",
  "`stringr`", "`str_ends`",
  "`stringr`", "`str_extract`",
  "`stringr`", "`str_length`",
  "`stringr`", "`str_remove`",
  "`stringr`", "`str_remove_all`",
  "`stringr`", "`str_replace`",
  "`stringr`", "`str_replace_all`",
  "`stringr`", "`str_starts`",
  "`stringr`", "`str_sub`",
  "`stringr`", "`str_to_lower`",
  "`stringr`", "`str_to_upper`",
) |> 
  arrange(Package)

rows <- seq_len(nrow(out) %/% 2)
kable(list(out[rows,], matrix(numeric(), nrow=0, ncol=1),
           out[-rows, ])) 
```

