---
title: "R and Polars expressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R and Polars expressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


When we use the `tidyverse`, we use R expressions in mainly three places: `filter()`,
`mutate()`, and `summarize()`.

```{r echo=FALSE}
library(dplyr, warn.conflicts = FALSE)
```


```{r eval=FALSE}
library(dplyr, warn.conflicts = FALSE)

filter(mtcars, am + gear > carb)
mutate(mtcars, x = (qsec - mean(qsec)) / sd(qsec))
mtcars |>
  group_by(cyl) |>
  summarize(x = mean(qsec) / sd(qsec))
```

This is very convenient but creates a challenge for `tidypolars`. While
it is possible to convert a Polars Data/LazyFrame to an R `data.frame`, apply
functions on it and then convert it back to Polars, it is strongly discouraged
to do so because it doesn't take advantage of Polars optimizations.

Indeed, Polars comes with dozens of built-in functions for maths (`median`,
`var`, `arccos`, ...), string manipulation (`len_chars`, `starts`, ...), and date-time
(`hour`, `quarter`, `ordinal_day`, ...). All of these functions are optimized
internally and are ran in parallel under the hood, which will not be the case if
we pass R functions.

However, using these Polars expressions would imply that we need to learn these
new functions and this new syntax. To avoid doing that, `tidypolars` will
automatically translate R expressions into Polars ones. Basically, **you can
keep writing R expressions in most situations**, and they will automatically be
translated to Polars syntax.

However, there are some situations where this might not work, so this vignette
explains the process and the limitations.

## How does `tidypolars` translate R expressions into Polars expressions?

When `tidypolars` receives an expression, it runs a function `translate()`
several times until all components are translated to their Polars equivalent.
There are four possible components: single values, column names, external objects,
and functions.


### Single values, column names, and external objects

If you pass a single value, like `x = 1` or `x = "a"`, it is wrapped into
`pl$lit()`. This is also the case for external objects with the difference that
these need to be wrapped in `{{ }}` and are evaluated before being wrapped into
`pl$lit()`.

Column names, like `x = mpg`, are wrapped into `pl$col()`.

```
x = "a"               ->  x = pl$lit("a")
x = {{ some_value }}  ->  x = pl$lit(*value*)
x = mpg               ->  x = pl$col("mpg")
```

### Functions


Functions are split into two categories: built-in functions (i.e functions
provided by base R or by other packages), and user-defined functions (UDF) that
are written by the user (you).


#### Built-in functions

In the first case, `tidypolars` checks the function name and whether it has
already been translated internally. For example, if we call the R function
`mean(x, trim = 2)`, then it looks for a translation of `mean()`. You can see the
list of supported R functions at the bottom of this vignette. Note that most of
essential base R functions are supported, as well as many functions from `dplyr`
or from `stringr` for example.

Now that `tidypolars` knows that a translation of `mean()` exists, it parses
the arguments in the call to translate them to the Polars syntax: internally,
`x` is converted to `pl$col("x")` if there is a column `"x"` in the data.
Sometimes, additional arguments do not have an equivalent in Polars. This is
the case for the argument `trim` here. In this case, `tidypolars` ignores
this argument and warns the user:

```{r}
library(tidypolars)
library(polars)

mtcars |>
  as_polars_df() |>
  mutate(x = mean(mpg, trim = 2))
```

This behavior [can be changed](https://tidypolars.etiennebacher.com/reference/tidypolars-options.html)
to throw an error instead.


#### User-defined functions

User-defined functions (UDF) are more challenging. Indeed, it is technically
possible to inspect the code inside a UDF, but rewriting it to match Polars syntax
would be extremely complicated. In this situation, you will have to rewrite your
custom function using Polars syntax so that it returns a Polars expression. For
example, we could make a function to standardize a column like this:

```{r}
pl_standardize <- function(x) {
  (x - x$mean()) / x$std()
}
```

Remember that the column name used as `x` will end up wrapped into `pl$col()`,
so to check that your function returns a Polars expression, you have to provide
a `pl$col()` call:

```{r}
pl_standardize(pl$col("mpg"))
```

This function correctly returns a Polars expression, so we can now use it like
any other function:

```{r}
mtcars |>
  as_polars_df() |>
  mutate(x = pl_standardize(mpg))
```


#### Special case: `across()`

[`across()`](https://dplyr.tidyverse.org/reference/across.html) is a very useful
function that applies a function (or a list of functions) to a selection of
columns. It accepts built-in functions, UDFs, and anonymous functions.

```{r}
mtcars |>
  as_polars_df() |>
  mutate(
    across(
      .cols = contains("a"),
      list(mean = mean, stand = pl_standardize, ~ sd(.x))
    )
  )
```

Similarly, UDFs and anonymous functions will error if they don't return a Polars
expression:

```{r error=TRUE}
mtcars |>
  as_polars_df() |>
  mutate(
    across(
      .cols = contains("a"),
      .fns = list(
        mean = mean,
        function(x) {
          (x - mean(x)) / sd(x)
        },
        ~ sd(.x)
      )
    )
  )
```

#### Non-translated functions

When a function provided in base R or in another package cannot be translated,
`tidypolars` usually throws an error (it's not necessary to understand what
`agrep()` does here, you should only know that it is not translated by
`tidypolars`):

```{r, error = TRUE}
dat <- pl$DataFrame(a = c("d", "e", "f"), foo = c(2, 1, 2))

dat |>
  filter(foo >= agrep("a", a))
```

This is because `foo >= agrep("a", a)` uses the values from the columns `a` and
`foo`, which are stored in the `polars` DataFrame. Therefore, `polars` needs a
translation of `agrep()`, which doesn't exist.

Now, let's assume that we have this data instead:

```{r}
dat <- pl$DataFrame(foo = c(2, 1, 2))
a <- c("d", "e", "f")

dat |>
  filter(foo >= agrep("a", a))
```

Starting from `tidypolars` 0.14.0, this doesn't error anymore. Why is that? The
reason is that while `foo >= agrep("a", a)` uses a column from the data (`foo`),
`agrep("a", a)` no longer does since `a` is an object in the environment and not
in the data anymore. Therefore, we can evaluate `agrep("a", a)` first and then use
its result when evaluating `foo >= agrep("a", a)`.

More generally, `tidypolars` checks whether a function uses columns from the
data and skips the translation part if it doesn't, allowing us to use more
functions than what is translated by `tidypolars` (only if they don't use
columns as inputs).

Note that expressions that don't use the data are evaluated before running
`polars` in the background so they don't benefit from `polars` parallel
evaluation for instance.

## A note on performance

As we saw above, a large part of the work of `tidypolars` is to translate
existing R code to its `polars` equivalent. This takes some time, especially
when translating R expressions used in `mutate()` or `filter()` for instance,
and it might be noticeable if you run `tidypolars` many times on small datasets.

This translation process takes more time as the number of expressions increases
but it doesn't depend on the size of the data. In other words, translating the
code from R to `polars` takes about the same time when applied on a dataset of
1M or 10M observations. Therefore, this drawback should become less and less
noticeable as the size of the dataset increases since most of the time will then
be taken by the actual computation performed by `polars`.
