---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We're going to use the dataset `who` provided in `tidyr` but we need to convert it to the `polars` format.

```{r setup}
library(polars)
library(tidypolars)

# who <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/tidyr/who.csv")
pl_who <- as_polars(read.csv("who.csv"))
```

The objective of `tidypolars` is to provide functions whose syntax is close to
the `tidyverse` while keeping the number of dependencies very low. Therefore, it
can be used in two ways depending on whether you already the `tidyverse` or not.

## You don't use the `tidyverse`

`tidypolars` provides two types of functions: some that replicate functions from 
the `tidyverse` and some that are "unique" to `polars`. The former are always
prefixed with `pl_`. For example, `tidypolars::pl_filter()` is equivalent to 
`dplyr::filter()`.

Like with the `tidyverse`, you can pipe multiple `tidypolars` functions:

```{r}
pl_who |> 
  pl_filter(year > 1990) |> 
  pl_drop_na(newrel_f3544) |> 
  pl_select(iso3, year, matches("^newrel(.*)_f"))
```

If you use `polars`' lazy API, you need to call `pl_collect()` at the end of the 
chained expression:

```{r}
pl_who_l <- as_polars(read.csv("who.csv"), lazy = TRUE)

out <- pl_who_l |> 
  pl_filter(year > 1990) |> 
  pl_drop_na(newrel_f3544) |> 
  pl_select(iso3, year, matches("^newrel(.*)_f"))

# this just prints the query
out

# this performs the query
out |> 
  pl_collect()
```

`tidypolars` always return a `DataFrame` or a `LazyFrame` so you can still use
`polars` original syntax after a chained expression:

```{r}
out <- pl_who |> 
  pl_filter(year > 1990) |> 
  pl_select(iso3, year, matches("^newrel(.*)_f"))

out

out$
  groupby("iso3", maintain_order = TRUE)$
  agg(
    pl$all()$mean()
  )$
  sort(c("iso3", "year"))
```

The code above only depends on `polars` and some `R` core packages.

## You already use the `tidyverse`

While the syntax above is very close to the `tidyverse` one, it can be tedious to
add the prefix `pl_` to every function. Therefore, if you already load `dplyr` or
other `tidyverse` packages in your code, you simply need to use `as_polars()` on 
your data (or read it with `polars` functions). The rest of the code doesn't need
to change at all. 

Here's an example of some `dplyr` code:

```{r}
library(dplyr, warn.conflicts = FALSE)

starwars |> 
  select(name, height, mass, species, sex) |> 
  mutate(
    BMI = mass / ((height/100) ^2)
  ) |> 
  select(BMI, everything()) |> 
  rename_with(.fn = toupper)
```

You can simply use a `polars` dataset instead, without changing anything (apart
from the argument name in `rename_with()`):

```{r}
library(tidypolars)

starwars |> 
  as_polars() |> 
  select(name, height, mass, species, sex) |> 
  mutate(
    BMI = mass / ((height/100) ^2)
  ) |> 
  select(BMI, everything()) |> 
  rename_with(fn = toupper)
```

**NOTE:** if you want to use the same functions as the `tidyverse` ones, you 
**must** load `dplyr` or the related packages first.
