---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
.custom_note {
  border: solid 3px #08505e;
  background-color: #0b788e;
  padding: 5px;
  margin-bottom: 10px;
  border-radius: 3px;
}
.custom_note > p {
  color: white
}
```

The first thing to do before using `tidypolars` is to get some data as a Polars
`DataFrame` or `LazyFrame`. You can read files with `polars::pl$read_*()`
functions (to import them as `DataFrame`s) or with `polars::pl$scan_*()` 
functions (to import them as `LazyFrame`s). `polars` can read various file 
formats, such as CSV, Parquet, or JSON.

You could also read data with other packages and then convert it with
`as_polars_df()` (of `as_polars_lf()` if you want to make it a 
`LazyFrame`). 

<div class="custom_note">
  <p><b>Note: </b><code>as_polars_df()</code> and <code>as_polars_lf()</code> are merely
  convenience functions to quickly convert data to a polars object, which is
  useful for showcase purposes. However, converting data from R to polars has
  some cost. In real-life usecases, be sure to load the data with the 
  <code>pl\$scan_\*()</code> or the <code>pl\$read_\*()</code> functions.</p>
</div>

Here, we're going to use the `who` dataset that is available in the `tidyr` 
package. I import it both as a classic R `data.frame` and as a Polars `DataFrame`
so that we can easily compare `dplyr` and `tidypolars` functions.

```{r setup}
library(polars)
library(tidypolars)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)

who_df <- tidyr::who
who_pl <- as_polars_df(tidyr::who)
```

`tidypolars` provides methods for `dplyr` and `tidyr` S3 generics. In simpler words, it 
means that you can use the same functions on a Polars `DataFrame` or `LazyFrame`
as in a classic `tidyverse` workflow and it should just work (if it doesn't,
please [open an issue](https://github.com/etiennebacher/tidypolars/issues)).
Note that you still need to load `dplyr` and `tidyr` in your code.

Here's an example of some `dplyr` and `tidyr` code on the classic R `data.frame`:

```{r}
who_df |> 
  filter(year > 1990) |> 
  drop_na(newrel_f3544) |> 
  select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  rename_with(.fn = toupper) |> 
  head()
```

We can simply use our Polars dataset instead:

```{r}
who_pl |> 
  filter(year > 1990) |> 
  drop_na(newrel_f3544) |> 
  select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  rename_with(.fn = toupper) |> 
  head()
```

If you use Polars lazy API, you need to call `compute()` at the end of the 
chained expression to evaluate the query:

```{r}
who_pl_lazy <- as_polars_lf(tidyr::who)

who_pl_lazy |> 
  filter(year > 1990) |> 
  drop_na(newrel_f3544) |> 
  select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  rename_with(.fn = toupper) |> 
  compute() |> 
  head()
```


`tidypolars` also supports many functions from `base`, `lubridate` or `stringr`.
When these are used inside `filter()`, `mutate()` or `summarize()`, `tidypolars`
will automatically convert them to use the Polars engine under the hood. Take
a look at the vignette ["R and Polars expressions"](https://tidypolars.etiennebacher.com/articles/r-and-polars-expressions) for more information.
