---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The first thing to do before using `tidypolars` is to get some data as a Polars
`DataFrame` or `LazyFrame`. You can read a CSV file with `polars::pl$read_csv()`
(to import it as a `DataFrame`) or with `polars::pl$scan_csv()` (to import it
as a `LazyFrame`).

You could also read the file with other packages and then convert it with
`as_polars()` (with the argument `lazy = TRUE` if you want to make it a 
`LazyFrame`). 

Here, we're going to use the `who` dataset that is available in the `tidyr` 
package. I import it both as a classic R `data.frame` and as a Polars `DataFrame`
so that we can easily compare `dplyr` and `tidypolars` functions.

```{r setup}
library(polars)
library(tidypolars)

link <- "https://vincentarelbundock.github.io/Rdatasets/csv/tidyr/who.csv"

who_df <- read.csv(link)
who_pl <- pl$read_csv(link, null_values = "NA")
```

The objective of `tidypolars` is to provide functions whose syntax is close to
the `tidyverse` while keeping the number of dependencies quite low. Therefore, it
can be used in two ways depending on whether you already the `tidyverse` or not.

## You already use the `tidyverse`

If you already load `dplyr` or other `tidyverse` packages in your code, you can
use the same code on a Polars `DataFrame` and it should work (if it doesn't,
feel free to [open an issue](https://github.com/etiennebacher/tidypolars/issues)).

Here's an example of some `dplyr` and `tidyr` code on the classic R `data.frame`:

```{r}
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)

who_df |> 
  filter(year > 1990) |> 
  drop_na(newrel_f3544) |> 
  select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  rename_with(.fn = toupper) |> 
  head()
```

We can simply use our Polars dataset instead:

```{r}
who_pl |> 
  filter(year > 1990) |> 
  drop_na(newrel_f3544) |> 
  select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  rename_with(.fn = toupper) |> 
  head()
```

If you use Polars lazy API, you need to call `collect()` at the end of the 
chained expression to evaluate the query:

```{r}
who_pl_lazy <- pl$scan_csv(link, null_values = "NA")

who_pl_lazy |> 
  filter(year > 1990) |> 
  drop_na(newrel_f3544) |> 
  select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  rename_with(.fn = toupper) |> 
  collect() |> 
  head()
```


**NOTE:** if you want to use the same functions as the `tidyverse` ones, you 
**must** load `dplyr` or the related packages (loading `tidypolars` alone is 
not enough).


## You don't want to use the `tidyverse`

As we saw above, it is extremely easy to use a `tidyverse` pipeline on some 
Polars data. However, maybe you would like to have access to this syntax without
importing `dplyr` or other packages (this could be the case if you want to use 
`tidypolars` inside a package for example). 

To do so, you can prefix the function calls with `pl_`, so that you would use
`pl_filter()` instead of `filter()` for instance. These functions have exactly
the same functionalities. Taking the example from above, you could have a script
where you only load `tidypolars` and run:

```{r}
who_pl |> 
  pl_filter(year > 1990) |> 
  pl_drop_na(newrel_f3544) |> 
  pl_select(iso3, year, matches("^newrel(.*)_f")) |> 
  arrange(iso3, year) |> 
  pl_rename_with(.fn = toupper) |> 
  head()
```

## About dependencies

If you care about dependencies (if you'd like to use `tidypolars` in a package 
for instance), `tidypolars` uses `polars` (no R package dependencies) and 
`tidyselect`, which is a low-level, widely tested package that has very rarely
breaking changes:

```{.r}
tools::package_dependencies("tidyselect", recursive = TRUE)
#> $tidyselect
#> [1] "cli"       "glue"      "lifecycle" "rlang"     "vctrs"     "withr"     "utils"     "methods"  
#> [9] "graphics"  "grDevices" "stats" 
```

